### 2014-08-06 Studies/2012_Chicago/Report/Chicago_Fire.tex
Bibliography: Graduate xfrac and cite packages to commoncommands
Chicago: update fds files, bibliography, and minor updates to report

% Extra packages
\usepackage{xfrac}
\usepackage{cite}

@@

Input data from various sources must be collected and documented to simulate a fire using FDS or any other fire model. For the simulation results presented in this document, information was obtained from two primary sources. The following information was gathered from the fire scene: the geometry of the building and the compartments being modeled, the size and location of exterior and interior ventilation openings, and documentation of fire damage to the building. The following information was gathered from witnesses, first responders, reports, and recorded media such as fire ground radios or videos: information on the timing of the fire development, the sequence and approximate timing of ventilation openings to the outside, and weather conditions at the time of the fire. The following subsections will describe the inputs that were used to construct the simulation, including any engineering approximations or assumptions that were made.

@@

%Fig.~(\ref{fig:smv_interior}) shows 3D Smokeview renderings of cutaways of the beta and delta sides of the structure interior.
%\begin{figure}[!ht]
%\begin{tabular*}{\textwidth}{l@{\extracolsep{\fill}}r}
%\includegraphics[width=.80\textwidth]{../Figures/interior_beta} \\
%\includegraphics[width=.80\textwidth]{../Figures/interior_delta} \\
%\end{tabular*}
%\caption{3D Smokeview rendering of structure interior looking from the charlie side to both the beta side (top) and to the delta side (bottom)}
%\label{fig:smv_interior}
%\end{figure}

@@

Wood fuel can be represented by the chemical formula, $\C\H_{1.7}\O_{0.74}\N_{0.002}$, with specified yields of carbon monoxide ($y_{\mathrm{CO}}=0.004 \; \mathrm{kg}/\mathrm{kg}$) and soot ($y_{\mathrm{C}}=0.015 \; \mathrm{kg}/\mathrm{kg}$). The product yields are expressed in terms of the amount of carbon monoxide or soot emitted per unit mass of fuel consumed (kg/kg) and can be found in the SFPE Handbook~\cite{SFPE:Tewarson}. A balanced chemical reaction for wood combustion can written as:

@@

The value for the heat of combustion of wood used in this simulation was~16,400 kJ/kg, based on data provided in the Society of Fire Protection Engineers (SFPE) Handbook~\cite{SFPE:Tewarson}. The heat of combustion quantifies the amount of energy per unit mass of the fuel. The following text defines Eq.~\ref{eq:wood_comb} for an FDS input file with the desired fuel and reaction properties discussed above:

@@

For the simulation, a measure of how well the flow field is resolved can be estimated by using the non-dimensional expression $D^*/\dx$. Here, $D^*$ is the characteristic fire diameter, $\dx$ is the nominal size of a mesh cell, and $\dQ$ is the total heat release rate of the fire:
\begin{equation}
D^* = \left(
     \frac{\dQ}{\rho_\infty \, c_p \, T_\infty \, \sqrt{g}}
     \right)^\frac{2}{5} 
\label{eq:mesh}
\end{equation}   
From the FDS User's Guide~\cite{FDS_Users_Guide}, the characteristic fire diameter is related to the characteristic fire size via the
relation $Q^* = (D^*/D)^{5/2}$. Here, $D$ is the physical diameter of the base of the fire specified in the simulation. Following Eq.~(\ref{eq:mesh}) and using a grid cell size of 10~cm, the characteristic fire diameter to cell size ($D^*/\dx$) ratio is 15.5 for the 3.3~MW porch fire and 23 for the 9~MW attic fire. Based on validation work performed for the U.S. Nuclear Regulatory Commission, $D^*/\dx$ values ranged between 4 and 16~\cite{NUREG_1824} and produced results that were judged to be adequate for engineering calculations. The grid resolution used in this model is within typical engineering $D^*/\dx$ values. As a result, all of the input model geometry lengths, ventilation openings (holes, doors, and windows), and fire areas will snap to the nearest 10~cm. Figure~\ref{fig:geom_grid} shows the front of the structure rendered in Smokeview with the 10 cm mesh.

@@

\begin{table}
\centering
\captionof{table}{Fire incident and simulation event timeline}\label{tab:firesim}
\begin{tabular}{ccl}
\toprule[1.5pt]
Incident Time  &  Simulation Time  &  Fire Behavior / Fireground Operation  \\
{[hh:mm:ss]}   &  {[s]}            &                                        \\
\midrule
\multirow{2}{*}{17:15:00} &    & \multirow{2}{*}{\parbox{9cm} {Fire is reported with heavy smoke coming from the front and rear of the structure's attic.}} \\ 
         & & \\[.25cm] %blank rows exist for wrapping text. # of blanks is 1 less than number of rows
\multirow{2}{*}{17:16:00} & \multirow{2}{*}{}   &  \multirow{2}{*}{\parbox{9cm} {Dispatch for a ``Still'' alarm - two engine companies, two truck companies, and a battalion chief.}} \\
         & & \\[.25cm]
\multirow{3}{*} {} &  \multirow{3}{*} {0} &  \multirow{3}{*}{\parbox{8cm} {FDS simulation starts. Attic fire ignites. Initial ventilation through rear of structure. Front door open.}} \\
         & & \\[0.25cm]
         & & \\[0.25cm]
\multirow{3}{*}{17:20:00} & \multirow{3}{*}{40}  & \multirow{3}{*}{\parbox{9cm} {Attic windows break - heavy smoke out of front attic window and heavy smoke and fire out of rear. Rear porch ignites in model.}} \\
         & & \\[.25cm]
         & & \\[.25cm]  
\multirow{2}{*}{}  & \multirow{2}{*}{45} & \multirow{2}{*}{\parbox{8cm}{Ventilation through wall penetration on rear of structure.}} \\
          & & \\[.25cm] 
\multirow{2}{*}{}  & \multirow{2}{*}{55}  & \multirow{2}{*}{\parbox{8cm}{Ventilation through wall penetration on left side of structure.}} \\ 
          & & \\[.25cm] 
\multirow{2}{*}{17:21:00} & \multirow{2}{*}{100} & \multirow{2}{*}{\parbox{8cm}{E123 captain and firefighter on second floor with 1\sfrac{3}{4}-inch hoseline.}} \\
         &  & \\[.25cm]
\multirow{2}{*}{17:21:30} & \multirow{2}{*}{130} &  \multirow{2}{*}{\parbox{8cm}{First floor exterior porch door opened. Porch fire ignites in FDS model.}} \\
        &  & \\[.25cm]
        & 135 & Additional ventilation through left side of structure. \\[.25cm] 
        & 160 & Top of interior door fails in FDS model. \\[.25cm] 
\multirow{3}{*}{17:23:00}    & \multirow{3}{*}{220} & \multirow{3}{*}{\parbox{8cm} {Heavy fire report in rear stairwell and enclosed porch. No response from E123 captain or firefighter.}} \\
         & & \\[.25cm]
         & & \\[.25cm]  
17:24:00    &   & Water applied to attic from the rear exterior position. \\[.25cm]
17:25:00    &   & SQ5 makes entry through front of structure. \\[.25cm]
\multirow{2}{*}{17:27:00}    & \multirow{2}{*}{300}  & \multirow{2}{*}{\parbox{8cm} {After hearing a Mayday call from SQ5, IC calls a Mayday. Simulation ends.}}\\
         & &  \\[.25cm]
17:29:00    &   & E123 captain carried out of structure. \\
\bottomrule[1.25pt]
\end{tabular}\par
\end{table}

@@

\section{Simulated Hallway Flow Path}
From on-scene reports (Section~\ref{fire_sum}) and the FDS simulation results (Section~\ref{temp}), the interior conditions on the second floor of the structure were initially tenable. From the NIOSH report~\cite{NIOSH:Bowyer}, while the IC was inside the structure on the second floor, only light haze and a glow around the door between the hallway and enclosed porch was observed. This indicated the presence of fire in the area of the enclosed porch. The simulation results indicate a high hazard area (temperatures greater than 260~$^{\circ}$C or 500~$^{\circ}$F) in the attic and the enclosed porch at this time. By the time the E123 captain and firefighter arrived on the second floor (Table~\ref{tab:firesim}) and approached the rear of the structure, the rear porch was fully involved. Temperature conditions greater than 260~$^{\circ}$C (500~$^{\circ}$F) were present on the porch side of the steel-faced, wood-framed door at the rear of the structure. Exposure to these conditions likely weakened the structural integrity of the door, which may have led to a partial failure. Figure~\ref{fig:chicago_doorfold} shows the state of the rear door after the fire. Note the large crease in the steel face of the door just above the knob.

@@

After the interior door failed, an interior flow path was created from the enclosed porch area into the kitchen of the structure. This resulted in a rapid change in the conditions within the flow path. The combustion gases at elevated temperature and pressure in the enclosed porch were vented by the failure of the door and flowed towards lower pressure areas in the hallways and kitchen. The arrows in Fig.~\ref{fig:flowpath_1} summarize the simulated flow direction of gases on the second floor following the door failure.

@@

A person is susceptible to second-degree burn injuries if exposed to temperatures greater than 55~$^{\circ}$C (130~$^{\circ}$F)~\cite{contactburn}. Although firefighters wear protective gear, gear only offers a finite amount of protection. The polycarbonate in a self contained breathing apparatus begins to soften when the material temperature reaches approximately 140~$^{\circ}$C (284~$^{\circ}$F)~\cite{mensch2011emergency}. Structural fire fighting coats and pants are tested to withstand temperatures of 260~$^{\circ}$C (500~$^{\circ}$F)~\cite{nfpa2013standard}. Prolonged exposure to elevated temperatures can result in a significant amount of heat transferred to the firefighter, putting him/her at risk. Therefore, for the analysis of the simulated exposure temperatures, the maximum visualized temperature is set to indicate all values that are greater than or equal to 260~$^{\circ}$C (500~$^{\circ}$F). This means that temperatures visualized at the maximum range in the displayed model output are at a minimum of 260~$^{\circ}$C (500~$^{\circ}$F); the actual simulated temperatures could in fact be greater.

@@

In this fire incident, the thermal degradation of the steel-faced, wood-framed door resulted in a rapid change in the thermal conditions in the hallway, which resulted in highly hazardous conditions.  Prediction of the failure mode and failure time of the door is not a reasonable expectation given the large number of unknowns that the firefighters faced on the fire ground.  In this case, the timing of the interior attack appears to have coincided with the failure time of the door. Research has been conducted~\cite{madrzykowski2009fire, kerber2009fire} and continues to be conducted by NIST, Underwriters Laboratories (UL), and others that demonstrates that applying water from the exterior into the fire area of a structure (prior to the start of interior operations) can significantly improve the safety of firefighters by reducing the thermal hazard from the fire and reducing the potential for developing high velocity hot gas flows within the structure.

>>>>

@@

Input data from various sources must be collected and documented to simulate a fire using FDS or any other fire model. For the simulation results presented in this document, information was obtained from two primary sources. The following information was gathered from the fire scene: the geometry of the building and the compartments being modeled, the size and location of exterior and interior ventilation openings, and documentation of fire damage to the building. The following information was gathered from witnesses, first responders, reports, and recorded media such as fire ground radios or videos: information on the timing of the fire development, the sequence and approximate timing of ventilation openings to the outside, and weather conditions at the time of the fire.

Based on the information collected, several engineering approximations were made. The actual ignition time is unknown.

The following subsections will describe the inputs that were used to construct the simulation, including any approximations or assumptions that were made.

@@

@@

Wood fuel can be represented by the chemical formula, $\C\H_{1.7}\O_{0.74}\N_{0.002}$, with specified yields of carbon monoxide ($y_{\mathrm{CO}}=0.004 \; \mathrm{kg}/\mathrm{kg}$) and soot ($y_{\mathrm{C}}=0.015 \; \mathrm{kg}/\mathrm{kg}$)~\cite{SFPE:Tewarson}. The product yields are expressed in terms of the amount of carbon monoxide or soot emitted per unit mass of fuel consumed (kg/kg) and can be found in the Society of Fire Protection Engineers (SFPE) Handbook~\cite{SFPE:Tewarson}. A balanced chemical reaction for wood combustion can written as:

@@

The value for the heat of combustion of wood used in this simulation was~16,400 kJ/kg, based on data provided in the SFPE Handbook~\cite{SFPE:Tewarson}. The heat of combustion quantifies the amount of energy per unit mass of the fuel. The following text defines Eq.~\ref{eq:wood_comb} for an FDS input file with the desired fuel and reaction properties discussed above:

@@

For the simulation, a measure of how well the flow field is resolved can be estimated by using the non-dimensional expression $D^*/\dx$. Here, $D^*$ is the characteristic fire diameter, $\dx$ is the nominal size of a mesh cell, and $\dQ$ is the total heat release rate of the fire:
\begin{equation}
D^* = \left(
     \frac{\dQ}{\rho_\infty \, c_p \, T_\infty \, \sqrt{g}}
     \right)^\frac{2}{5} 
\label{eq:mesh}
\end{equation}   
From the FDS User's Guide~\cite{FDS_Users_Guide}, the characteristic fire diameter is related to the characteristic fire size via the
relation $Q^* = (D^*/D)^{5/2}$. Here, $D$ is the physical diameter of the base of the fire specified in the simulation. Following Eq.~(\ref{eq:mesh}) and using a grid cell size of 10~cm, the characteristic fire diameter to cell size ($D^*/\dx$) ratio is 15.5 for the 3.3~MW porch fire and 23 for the 9~MW attic fire. Based on validation work performed for the U.S. Nuclear Regulatory Commission, $D^*/\dx$ values ranged between 4 and 16~\cite{NUREG_1824} and produced results that were judged to be adequate for engineering calculations. The grid resolution used in this model is within  or exceeds typical engineering $D^*/\dx$ values. As a result, all of the input model geometry lengths, ventilation openings (holes, doors, and windows), and fire areas will snap to the nearest 10~cm. Figure~\ref{fig:geom_grid} shows the front of the structure rendered in Smokeview with the 10 cm mesh.

@@

\begin{table}
\centering
\captionof{table}{Fire incident and simulation event timeline}\label{tab:firesim}
\begin{tabular}{ccl}
\toprule[1.5pt]
Incident Time  &  Simulation Time  &  Fire Behavior / Fireground Operation  \\
{[hh:mm:ss]}   &  {[s]}            &                                        \\
\midrule
\multirow{2}{*}{17:15:00} &    & \multirow{2}{*}{\parbox{9cm} {Fire is reported with heavy smoke coming from the front and rear of the structure's attic.}} \\ 
         & & \\[.25cm] %blank rows exist for wrapping text. # of blanks is 1 less than number of rows
\multirow{2}{*}{17:16:00} & \multirow{2}{*}{}   &  \multirow{2}{*}{\parbox{9cm} {Dispatch for a ``Still'' alarm - two engine companies, two truck companies, and a battalion chief.}} \\
         & & \\[.25cm]
\multirow{3}{*} {} &  \multirow{3}{*} {0} &  \multirow{3}{*}{\parbox{8cm} {FDS simulation starts. Attic fire ignites. Initial ventilation through rear of structure. Front door open.}} \\
         & & \\[0.25cm]
         & & \\[0.25cm]
\multirow{3}{*}{17:20:00} & \multirow{3}{*}{40}  & \multirow{3}{*}{\parbox{9cm} {Attic windows break - heavy smoke out of front attic window and heavy smoke and fire out of rear. Rear porch ignites in model.}} \\
         & & \\[.25cm]
         & & \\[.25cm]  
\multirow{2}{*}{}  & \multirow{2}{*}{45} & \multirow{2}{*}{\parbox{8cm}{Ventilation through wall penetration on rear of structure.}} \\
          & & \\[.25cm] 
\multirow{2}{*}{}  & \multirow{2}{*}{55}  & \multirow{2}{*}{\parbox{8cm}{Ventilation through wall penetration on left side of structure.}} \\ 
          & & \\[.25cm] 
\multirow{2}{*}{17:21:00} & \multirow{2}{*}{100} & \multirow{2}{*}{\parbox{8cm}{E123 captain and firefighter on second floor with 1\sfrac{3}{4}-inch hoseline.}} \\
         &  & \\[.25cm]
\multirow{2}{*}{17:21:30} & \multirow{2}{*}{130} &  \multirow{2}{*}{\parbox{8cm}{First floor exterior porch door opened. Porch fire ignites in FDS model.}} \\
        &  & \\[.25cm]
        & 135 & Ventilation through left side of structure. \\[.25cm] 
        & 160 & Top of interior door fails in FDS model. \\[.25cm] 
\multirow{3}{*}{17:23:00}    & \multirow{3}{*}{220} & \multirow{3}{*}{\parbox{8cm} {Heavy fire report in rear stairwell and enclosed porch. No response from E123 captain or firefighter.}} \\
         & & \\[.25cm]
         & & \\[.25cm]  
17:24:00    &   & Water applied to attic from the rear exterior position. \\[.25cm]
17:25:00    &   & SQ5 makes entry through front of structure. \\[.25cm]
\multirow{2}{*}{17:27:00}    & \multirow{2}{*}{300}  & \multirow{2}{*}{\parbox{8cm} {After hearing a Mayday call from SQ5, IC calls a Mayday. Simulation ends.}}\\
         & &  \\[.25cm]
17:29:00    &   & E123 captain carried out of structure. \\
\bottomrule[1.25pt]
\end{tabular}\par
For specific information regarding ventilation areas used in FDS model, refer to Table~\ref{tab:vents}.
\end{table}

@@

\section{Simulated Hallway Flow Path}
From on-scene reports (Section~\ref{fire_sum}) and the FDS simulation results (Section~\ref{temp}), the interior conditions on the second floor of the structure were initially tenable. From the NIOSH report~\cite{NIOSH:Bowyer}, while the IC was inside the structure on the second floor, only light haze and a glow around the door between the hallway and enclosed porch was observed. This indicated the presence of fire in the area of the enclosed porch. The simulation results indicate a high hazard area (temperatures greater than 260~$^{\circ}$C or 500~$^{\circ}$F) in the attic and the enclosed porch at this time. By the time the E123 captain and firefighter arrived on the second floor (Table~\ref{tab:firesim}) and approached the rear of the structure, the rear porch was fully involved. Temperature conditions greater than 260~$^{\circ}$C (500~$^{\circ}$F) were present on the porch side of the steel-faced, wood-framed door at the rear of the structure. Exposure to these conditions likely weakened the structural integrity of the door, which may have led to a partial failure (Table~\ref{tab:firesim}). Figure~\ref{fig:chicago_doorfold} shows the state of the rear door after the fire. Note the large crease in the steel face of the door just above the knob.

@@

After the interior door failed, an interior flow path was created from the enclosed porch area into the kitchen of the structure. This resulted in a rapid change in the conditions within the flow path. The combustion gases at elevated temperature and pressure in the enclosed porch were vented by the failure of the door and flowed towards lower pressure areas in the hallways and kitchen. The arrows in Fig.~\ref{fig:flowpath_1} summarize the simulated flow direction of gases on the second floor following the door failure. Red colored arrows indicate that the gases were at an elevated temperature, while the blue arrows indicate that the gases were at temperatures close to ambient.  The arrow that transitions from red to blue indicates that the gas began to cool as it approached the front of the structure. Note that at the second story doorway towards the front of the structure there are two arrows; this is an indication of bi-directional flow. The cool ambient air flowed in at the bottom while the gas at elevated temperature flowed out the top.

@@

A person is susceptible to second-degree burn injuries if exposed to temperatures greater than 55~$^{\circ}$C (130~$^{\circ}$F)~\cite{contactburn}. Although firefighters wear protective gear, gear only offers a finite amount of protection. The polycarbonate in a self contained breathing apparatus begins to soften when the material temperature reaches approximately 140~$^{\circ}$C (284~$^{\circ}$F)~\cite{mensch2011emergency}. Structural fire fighting coats and pants are tested to withstand temperatures of 260~$^{\circ}$C (500~$^{\circ}$F)~\cite{nfpa2013standard}. Prolonged exposure to elevated temperatures can result in a significant amount of heat transferred to the firefighter, putting him/her at risk. Exposure of equipment to temperatures of 260~$^{\circ}$C (500~$^{\circ}$F), represents a Class III exposure~\cite{Donnelly2006}. Firefighters face increased risk when facing Class III exposures for greater than five minutes~\cite{Donnelly2006}.  Therefore, for the analysis of the simulated exposure temperatures, the maximum visualized temperature is set to indicate all values that are greater than or equal to 260~$^{\circ}$C (500~$^{\circ}$F). This means that temperatures visualized at the maximum range in the displayed model output are at a minimum of 260~$^{\circ}$C (500~$^{\circ}$F); the actual simulated temperatures could in fact be greater.

@@

In this fire incident, the thermal degradation of the steel-faced, wood-framed door resulted in a rapid change in the thermal conditions in the hallway, which resulted in highly hazardous conditions. The these conditions would be equivalent to or greater than a Class III exposure~\cite{Donnelly2006}. Prediction of the failure mode and failure time of the door is not a reasonable expectation given the large number of unknowns that the firefighters faced on the fire ground.  In this case, the timing of the interior attack appears to have coincided with the failure time of the door. Research has been conducted~\cite{madrzykowski2009fire, kerber2009fire} and continues to be conducted by NIST, Underwriters Laboratories (UL), and others that demonstrates that applying water from the exterior into the fire area of a structure (prior to the start of interior operations) can significantly improve the safety of firefighters by reducing the thermal hazard from the fire and reducing the potential for developing high velocity hot gas flows within the structure.
